### 2025-06-21 modern-portfolio-theory.qmd
20250621 - efficient frontier

@@

@@

@@

plot(
  efficientFrontierData$volatility,
  efficientFrontierData$return,
  pch = 19,
  cex = 0.5,
  type = "o",
  xlab = "Standard Deviation of Projected Fantasy Points Across Sources",
  ylab = "Projected Fantasy Points")
```

@@

@@

@@

@@

plot(
  efficientFrontierData2$volatility,
  efficientFrontierData2$return,
  pch = 19,
  cex = 0.5,
  type = "o",
  xlab = "Standard Deviation of Fantasy Points Across Weeks",
  ylab = "Weekly Fantasy Points")
```

>>>>

As shown below, the correlation between mean and standard deviation of projected points is positive.
That is, greater expected returns (i.e., mean of projected fantasy points) are associated with greater risk or volatility (i.e., standard deviation of projected fantasy points).
Thus, to obtain a lineup that scores greater fantasy points, it may be necessary to take on greater risk.
Moreover, you would not want take take on more risk for the same number of expected fantasy points; in addition, you would not want to score fewer fantasy points for the same amount of risk [@Hitchings2012].

```{r}
cor.test(
  ~ mean + sd,
  data = all_proj_summary
)
```

@@

all_proj_summary %>% 
  filter(position == "QB" & mean > 300) %>% 
  arrange(-sd)

all_proj_summary %>% 
  filter(position == "QB" & mean > 150) %>% 
  arrange(sd)

all_proj_summary %>% 
  filter(position == "RB" & mean > 200) %>% 
  arrange(-sd)

all_proj_summary %>% 
  filter(position == "RB" & mean > 100) %>% 
  arrange(sd)

all_proj_summary %>% 
  filter(position == "WR" & mean > 200) %>% 
  arrange(-sd)

all_proj_summary %>% 
  filter(position == "WR" & mean > 100) %>% 
  arrange(sd)
```

```{r}
hiLo_permutations <- expand.grid(
  rep(list(c("Hi", "Lo")), 3),
  stringsAsFactors = FALSE
)

qbRbWr_projections <- data.frame(
  condition = c("Hi", "Lo"),
  qb_name = c("Josh Allen", "Matthew Stafford"),
  rb_name = c("Saquon Barkley", "Joe Mixon"),
  wr_name = c("Ja'Marr Chase", "Courtland Sutton")
)

qbRbWr_projections$qb_mean <- c(
  all_proj_summary$mean[which(all_proj_summary$position == "QB" & all_proj_summary$name == qbRbWr_projections$qb_name[1])],
  all_proj_summary$mean[which(all_proj_summary$position == "QB" & all_proj_summary$name == qbRbWr_projections$qb_name[2])])

qbRbWr_projections$qb_sd <- c(
  all_proj_summary$sd[which(all_proj_summary$position == "QB" & all_proj_summary$name == qbRbWr_projections$qb_name[1])],
  all_proj_summary$sd[which(all_proj_summary$position == "QB" & all_proj_summary$name == qbRbWr_projections$qb_name[2])])

qbRbWr_projections$rb_mean <- c(
  all_proj_summary$mean[which(all_proj_summary$position == "RB" & all_proj_summary$name == qbRbWr_projections$rb_name[1])],
  all_proj_summary$mean[which(all_proj_summary$position == "RB" & all_proj_summary$name == qbRbWr_projections$rb_name[2])])

qbRbWr_projections$rb_sd <- c(
  all_proj_summary$sd[which(all_proj_summary$position == "RB" & all_proj_summary$name == qbRbWr_projections$rb_name[1])],
  all_proj_summary$sd[which(all_proj_summary$position == "RB" & all_proj_summary$name == qbRbWr_projections$rb_name[2])])

qbRbWr_projections$wr_mean <- c(
  all_proj_summary$mean[which(all_proj_summary$position == "WR" & all_proj_summary$name == qbRbWr_projections$wr_name[1])],
  all_proj_summary$mean[which(all_proj_summary$position == "WR" & all_proj_summary$name == qbRbWr_projections$wr_name[2])])

qbRbWr_projections$wr_sd <- c(
  all_proj_summary$sd[which(all_proj_summary$position == "WR" & all_proj_summary$name == qbRbWr_projections$wr_name[1])],
  all_proj_summary$sd[which(all_proj_summary$position == "WR" & all_proj_summary$name == qbRbWr_projections$wr_name[2])])

qbRbWr_projections$qb_lastName <- sapply(strsplit(qbRbWr_projections$qb_name, " "), function(x) tail(x, 1))
qbRbWr_projections$rb_lastName <- sapply(strsplit(qbRbWr_projections$rb_name, " "), function(x) tail(x, 1))
qbRbWr_projections$wr_lastName <- sapply(strsplit(qbRbWr_projections$wr_name, " "), function(x) tail(x, 1))

team_projections <- data.frame(
  team = apply(hiLo_permutations, 1, paste0, collapse = ""),
  team_name = NA,
  team_mean = NA,
  team_sd = NA,
  qb = hiLo_permutations$Var1,
  rb = hiLo_permutations$Var2,
  wr = hiLo_permutations$Var3,
  qb_name = NA,
  qb_lastName = NA,
  qb_mean = NA,
  qb_sd = NA,
  rb_name = NA,
  rb_lastName = NA,
  rb_mean = NA,
  rb_sd = NA,
  wr_name = NA,
  wr_lastName = NA,
  wr_mean = NA,
  wr_sd = NA
)

team_projections$qb_name[which(team_projections$qb == "Hi")] <- qbRbWr_projections$qb_name[which(qbRbWr_projections$condition == "Hi")]
team_projections$qb_name[which(team_projections$qb == "Lo")] <- qbRbWr_projections$qb_name[which(qbRbWr_projections$condition == "Lo")]
team_projections$qb_lastName[which(team_projections$qb == "Hi")] <- qbRbWr_projections$qb_lastName[which(qbRbWr_projections$condition == "Hi")]
team_projections$qb_lastName[which(team_projections$qb == "Lo")] <- qbRbWr_projections$qb_lastName[which(qbRbWr_projections$condition == "Lo")]
team_projections$qb_mean[which(team_projections$qb == "Hi")] <- qbRbWr_projections$qb_mean[which(qbRbWr_projections$condition == "Hi")]
team_projections$qb_mean[which(team_projections$qb == "Lo")] <- qbRbWr_projections$qb_mean[which(qbRbWr_projections$condition == "Lo")]
team_projections$qb_sd[which(team_projections$qb == "Hi")] <- qbRbWr_projections$qb_sd[which(qbRbWr_projections$condition == "Hi")]
team_projections$qb_sd[which(team_projections$qb == "Lo")] <- qbRbWr_projections$qb_sd[which(qbRbWr_projections$condition == "Lo")]

team_projections$rb_name[which(team_projections$rb == "Hi")] <- qbRbWr_projections$rb_name[which(qbRbWr_projections$condition == "Hi")]
team_projections$rb_name[which(team_projections$rb == "Lo")] <- qbRbWr_projections$rb_name[which(qbRbWr_projections$condition == "Lo")]
team_projections$rb_lastName[which(team_projections$rb == "Hi")] <- qbRbWr_projections$rb_lastName[which(qbRbWr_projections$condition == "Hi")]
team_projections$rb_lastName[which(team_projections$rb == "Lo")] <- qbRbWr_projections$rb_lastName[which(qbRbWr_projections$condition == "Lo")]
team_projections$rb_mean[which(team_projections$rb == "Hi")] <- qbRbWr_projections$rb_mean[which(qbRbWr_projections$condition == "Hi")]
team_projections$rb_mean[which(team_projections$rb == "Lo")] <- qbRbWr_projections$rb_mean[which(qbRbWr_projections$condition == "Lo")]
team_projections$rb_sd[which(team_projections$rb == "Hi")] <- qbRbWr_projections$rb_sd[which(qbRbWr_projections$condition == "Hi")]
team_projections$rb_sd[which(team_projections$rb == "Lo")] <- qbRbWr_projections$rb_sd[which(qbRbWr_projections$condition == "Lo")]

team_projections$wr_name[which(team_projections$wr == "Hi")] <- qbRbWr_projections$wr_name[which(qbRbWr_projections$condition == "Hi")]
team_projections$wr_name[which(team_projections$wr == "Lo")] <- qbRbWr_projections$wr_name[which(qbRbWr_projections$condition == "Lo")]
team_projections$wr_lastName[which(team_projections$wr == "Hi")] <- qbRbWr_projections$wr_lastName[which(qbRbWr_projections$condition == "Hi")]
team_projections$wr_lastName[which(team_projections$wr == "Lo")] <- qbRbWr_projections$wr_lastName[which(qbRbWr_projections$condition == "Lo")]
team_projections$wr_mean[which(team_projections$wr == "Hi")] <- qbRbWr_projections$wr_mean[which(qbRbWr_projections$condition == "Hi")]
team_projections$wr_mean[which(team_projections$wr == "Lo")] <- qbRbWr_projections$wr_mean[which(qbRbWr_projections$condition == "Lo")]
team_projections$wr_sd[which(team_projections$wr == "Hi")] <- qbRbWr_projections$wr_sd[which(qbRbWr_projections$condition == "Hi")]
team_projections$wr_sd[which(team_projections$wr == "Lo")] <- qbRbWr_projections$wr_sd[which(qbRbWr_projections$condition == "Lo")]

team_projections$team_name <- apply(
  team_projections[, c("qb_lastName", "rb_lastName", "wr_lastName")],
  1,
  paste,
  collapse = "/"
)

team_projections$team_mean <- rowSums(team_projections[,c("qb_mean","rb_mean","wr_mean")])
team_projections$team_sd <- rowSums(team_projections[,c("qb_sd","rb_sd","wr_sd")])
```

In @fig-teamPortfolio1, we depict the combined mean of projected fantasy points for a Quarterback/Running Back/Wide Receiver combination as a function of the standard deviation of their projected fantasy points across sources.

```{r}
#| label: fig-teamPortfolio1
#| fig-cap: "Mean of Projected Fantasy Points For a Quarterback/Running Back/Wide Receiver Combination as a Function of the Standard Deviation of the Projected Fantasy Points Across Sources."
#| fig-alt: "Mean of Projected Fantasy Points For a Quarterback/Running Back/Wide Receiver Combination as a Function of the Standard Deviation of the Projected Fantasy Points Across Sources."

ggplot2::ggplot(
  data = team_projections,
  aes(
    x = team_sd,
    y = team_mean,
    label = team_name)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = team_name)) +
  #geom_text(
  #  hjust = 0,
  #  vjust = 0) +
  #geom_smooth() +
  coord_cartesian(
    xlim = c(NA,NA),
    ylim = c(NA,NA),
    expand = TRUE) +
  labs(
    x = "Standard Deviation of Players' Projected Fantasy Points (Season) Across Sources",
    y = "Mean of Players' Projected Fantasy Points (Season) Across Sources",
    title = "Mean vs Standard Deviation of Players' Projected Fantasy Points"
  ) +
  theme_classic()
```

@@

When you evaluate all possible combinations of players, you can obtain the expected returns and standard deviations for each player combination.
Based on that, you can determine the maximum number of fantasy points (i.e., the best possible team) at each level of risk.
The best expected returns at each level of risk is known as the efficient frontier.
The efficient frontier of projected fantasy points across sources is in @fig-efficientFrontierAcrossProjections.

@@

plot(
  efficientFrontierData$volatility,
  efficientFrontierData$return,
  pch = 19,
  cex = 0.5,
  type = "o",
  xlab = "Standard Deviation of Projected Fantasy Points Across Sources",
  ylab = "Projected Fantasy Points",
  main = "Efficient Frontier of Projected Fantasy Points Across Sources")
```

@@

```{r}
cor.test(
  ~ mean + sd,
  data = player_stats_weekly_recent_summary
)
```

@@

player_stats_weekly_recent_summary %>% 
  filter(position == "QB" & mean > 20) %>% 
  arrange(-sd)

player_stats_weekly_recent_summary %>% 
  filter(position == "QB" & mean > 10) %>% 
  arrange(sd)

player_stats_weekly_recent_summary %>% 
  filter(position == "RB" & mean > 17) %>% 
  arrange(-sd)

player_stats_weekly_recent_summary %>% 
  filter(position == "RB" & mean > 10) %>% 
  arrange(sd)

player_stats_weekly_recent_summary %>% 
  filter(position == "WR" & mean > 17) %>% 
  arrange(-sd)

player_stats_weekly_recent_summary %>% 
  filter(position == "WR" & mean > 10) %>% 
  arrange(sd)
```

```{r}
qbRbWr_projections2 <- data.frame(
  condition = c("Hi", "Lo"),
  qb_name = c("Josh Allen", "Patrick Mahomes"),
  rb_name = c("Saquon Barkley", "David Montgomery"),
  wr_name = c("Ja'Marr Chase", "Stefon Diggs")
)

qbRbWr_projections2$qb_mean <- c(
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "QB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$qb_name[1])],
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "QB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$qb_name[2])])

qbRbWr_projections2$qb_sd <- c(
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "QB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$qb_name[1])],
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "QB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$qb_name[2])])

qbRbWr_projections2$rb_mean <- c(
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "RB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$rb_name[1])],
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "RB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$rb_name[2])])

qbRbWr_projections2$rb_sd <- c(
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "RB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$rb_name[1])],
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "RB" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$rb_name[2])])

qbRbWr_projections2$wr_mean <- c(
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "WR" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$wr_name[1])],
  player_stats_weekly_recent_summary$mean[which(player_stats_weekly_recent_summary$position == "WR" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$wr_name[2])])

qbRbWr_projections2$wr_sd <- c(
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "WR" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$wr_name[1])],
  player_stats_weekly_recent_summary$sd[which(player_stats_weekly_recent_summary$position == "WR" & player_stats_weekly_recent_summary$name == qbRbWr_projections2$wr_name[2])])

qbRbWr_projections2$qb_lastName <- sapply(strsplit(qbRbWr_projections2$qb_name, " "), function(x) tail(x, 1))
qbRbWr_projections2$rb_lastName <- sapply(strsplit(qbRbWr_projections2$rb_name, " "), function(x) tail(x, 1))
qbRbWr_projections2$wr_lastName <- sapply(strsplit(qbRbWr_projections2$wr_name, " "), function(x) tail(x, 1))

team_projections2 <- data.frame(
  team = apply(hiLo_permutations, 1, paste0, collapse = ""),
  team_name = NA,
  team_mean = NA,
  team_sd = NA,
  qb = hiLo_permutations$Var1,
  rb = hiLo_permutations$Var2,
  wr = hiLo_permutations$Var3,
  qb_name = NA,
  qb_lastName = NA,
  qb_mean = NA,
  qb_sd = NA,
  rb_name = NA,
  rb_lastName = NA,
  rb_mean = NA,
  rb_sd = NA,
  wr_name = NA,
  wr_lastName = NA,
  wr_mean = NA,
  wr_sd = NA
)

team_projections2$qb_name[which(team_projections2$qb == "Hi")] <- qbRbWr_projections2$qb_name[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$qb_name[which(team_projections2$qb == "Lo")] <- qbRbWr_projections2$qb_name[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$qb_lastName[which(team_projections2$qb == "Hi")] <- qbRbWr_projections2$qb_lastName[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$qb_lastName[which(team_projections2$qb == "Lo")] <- qbRbWr_projections2$qb_lastName[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$qb_mean[which(team_projections2$qb == "Hi")] <- qbRbWr_projections2$qb_mean[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$qb_mean[which(team_projections2$qb == "Lo")] <- qbRbWr_projections2$qb_mean[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$qb_sd[which(team_projections2$qb == "Hi")] <- qbRbWr_projections2$qb_sd[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$qb_sd[which(team_projections2$qb == "Lo")] <- qbRbWr_projections2$qb_sd[which(qbRbWr_projections2$condition == "Lo")]

team_projections2$rb_name[which(team_projections2$rb == "Hi")] <- qbRbWr_projections2$rb_name[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$rb_name[which(team_projections2$rb == "Lo")] <- qbRbWr_projections2$rb_name[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$rb_lastName[which(team_projections2$rb == "Hi")] <- qbRbWr_projections2$rb_lastName[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$rb_lastName[which(team_projections2$rb == "Lo")] <- qbRbWr_projections2$rb_lastName[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$rb_mean[which(team_projections2$rb == "Hi")] <- qbRbWr_projections2$rb_mean[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$rb_mean[which(team_projections2$rb == "Lo")] <- qbRbWr_projections2$rb_mean[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$rb_sd[which(team_projections2$rb == "Hi")] <- qbRbWr_projections2$rb_sd[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$rb_sd[which(team_projections2$rb == "Lo")] <- qbRbWr_projections2$rb_sd[which(qbRbWr_projections2$condition == "Lo")]

team_projections2$wr_name[which(team_projections2$wr == "Hi")] <- qbRbWr_projections2$wr_name[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$wr_name[which(team_projections2$wr == "Lo")] <- qbRbWr_projections2$wr_name[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$wr_lastName[which(team_projections2$wr == "Hi")] <- qbRbWr_projections2$wr_lastName[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$wr_lastName[which(team_projections2$wr == "Lo")] <- qbRbWr_projections2$wr_lastName[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$wr_mean[which(team_projections2$wr == "Hi")] <- qbRbWr_projections2$wr_mean[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$wr_mean[which(team_projections2$wr == "Lo")] <- qbRbWr_projections2$wr_mean[which(qbRbWr_projections2$condition == "Lo")]
team_projections2$wr_sd[which(team_projections2$wr == "Hi")] <- qbRbWr_projections2$wr_sd[which(qbRbWr_projections2$condition == "Hi")]
team_projections2$wr_sd[which(team_projections2$wr == "Lo")] <- qbRbWr_projections2$wr_sd[which(qbRbWr_projections2$condition == "Lo")]

team_projections2$team_name <- apply(
  team_projections2[, c("qb_lastName", "rb_lastName", "wr_lastName")],
  1,
  paste,
  collapse = "/"
)

team_projections2$team_mean <- rowSums(team_projections2[,c("qb_mean","rb_mean","wr_mean")])
team_projections2$team_sd <- rowSums(team_projections2[,c("qb_sd","rb_sd","wr_sd")])
```

In @fig-teamPortfolio2, we depict the combined mean of projected fantasy points for a Quarterback/Running Back/Wide Receiver combination as a function of the standard deviation of their projected fantasy points across sources.

```{r}
#| label: fig-teamPortfolio2
#| fig-cap: "Mean of Weekly Fantasy Points For a Quarterback/Running Back/Wide Receiver Combination as a Function of the Week-to-Week Standard Deviation of their Fantasy Points."
#| fig-alt: "Mean of Weekly Fantasy Points For a Quarterback/Running Back/Wide Receiver Combination as a Function of the Week-to-Week Standard Deviation of their Fantasy Points."

ggplot2::ggplot(
  data = team_projections2,
  aes(
    x = team_sd,
    y = team_mean,
    label = team_name)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = team_name)) +
  #geom_text(
  #  hjust = 0,
  #  vjust = 0) +
  #geom_smooth() +
  coord_cartesian(
    xlim = c(NA,NA),
    ylim = c(NA,NA),
    expand = TRUE) +
  labs(
    x = "Standard Deviation of Players' Projected Fantasy Points (Season) Across Sources",
    y = "Mean of Players' Projected Fantasy Points (Season) Across Sources",
    title = "Mean vs Standard Deviation of Players' Projected Fantasy Points"
  ) +
  theme_classic()
```

@@

The efficient frontier of fantasy points across weeks is in @fig-efficientFrontierAcrossWeeks.

@@

plot(
  efficientFrontierData2$volatility,
  efficientFrontierData2$return,
  pch = 19,
  cex = 0.5,
  type = "o",
  xlab = "Standard Deviation of Fantasy Points Across Weeks",
  ylab = "Weekly Fantasy Points",
  main = "Efficient Frontier of Fantasy Points Across Weeks")
```
